The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-11-05 script/en/chapters/paper_and_pencil.tex
CT script: Many small changes; especially referenced CT2, JCT and Sage where previously only CT1 was mentioned.

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Writing rule: Capitel header with all nouns/verbs-words capitalized,
%               All other headers in the normal lower/upper case manner
%               like sentences (without a dot at the end).
% Writing rule: Use American English instead of British English consistently.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% Update and corrections: B. Esslinger, June 2005
% Update B. Esslinger and Minh Van Nguyen, November 2009 
%------------------------------------------------------------------------------

@@

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005; Updates Minh Van Nguyen and B. Esslinger, November 2009 and June 2010)
\index{Paper and pencil methods}

\begin{center}
\fbox{\parbox{15cm}{%
{\em Edgar Allan Poe\index{Poe, Edgar Allan}: 
A Few Words on Secret Writing, 1841}\\
Few persons can be made to believe that it is not quite an easy thing
to invent a method of secret writing which shall baffle investigation.
Yet it may be roundly asserted that human ingenuity cannot concoct a
cipher which human ingenuity cannot resolve.}}
\end{center}

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes to this chapter describe how the methods can be
performed using CrypTool 1 (CT1)\index{CT1}, CrypTool 2 (CT2)\index{CT2}
or JCrypTool (JCT)\index{JCT}.
\index{CT1} \index{CT2} \index{JCT}
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code using the computer algebra system Sage\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\item {\bf Rail Fence}\footnote{%
   This method can directly be found in CrypTool\index{CrypTool} at the menu item
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Scytale / Rail Fence}.
   You can simulate this method also under the menu {\bf Encrypt/Decrypt \textbackslash{} 
   Symmetric (classic) \textbackslash{} Permutation}: For a Rail Fence with
   2 lines use as key ``B,A'' and accept the default settings (only one
   permutation, where your input is done line-by-line and the output is
   taken column-by-column). 
   Using the key ``A,B'' would start the zigzag pattern below in the
   way, that the first letter is written into the first line instead of the
   second line.}
   \cite{pp:Singh1999}\index{Rail Fence cipher}:
   The characters of a message are alternately written in two (or more) lines,
   creating a zigzag pattern. The resulting ciphertext is read out 
   line by line.\\
   This is more a children's method.

   Plaintext\footnote{If the alphabet only uses 26 letters, we write the
   plaintext in small letters and the ciphertext in capital letters.}%
   : an example of transposition

@@

\item {\bf Scytale}\footnote{%
   This method can directly be found in CrypTool\index{CrypTool} at the menu item
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Scytale / Rail Fence}.
   As this method is a special case of a simple columnar transposition, you also can
   simulate it in CrypTool\index{CrypTool} under the 
   menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
   Permutation}: For the Scytale within the dialog box only the first 
   permutation is used. If the wood has e.g.\ 4 angles use as key ``1,2,3,4''.
   This is equivalent to write the text horizontally in blocks of 4 letters 
   in a matrix and to read it out vertically . 
   Because the key is in an in ascending order, the Scytale is denoted as
   an identical permutation. And because writing and read-out is done only
   once it is a simple (and no double) permutation.}
   \cite{pp:Singh1999}\index{Scytale}%
   : 
   This method was probably used since 600 B.C. -- a description
   of how it operated is not known from before Plutarch (50-120 B.C.).\\
   A long strip of paper is wrapped around a wooden cylinder and then the 
   message is written along the length of this strip. The ciphertext is 
   produced by unwinding the strip.

@@

   \hypertarget{turning-grille}{}
\item {\bf Turning grille} \cite{pp:Savard1999}%
   \index{Turning grille}: 
   The German army used turning 
   grilles during WW1\footnote{The turning grille was already invented in 
   1881 by Eduard Fleissner von Wostrowitz.\\
   A good visualization can be found under www.turning-grille.com.}%
   . 
   A square grille serves as a stencil, a quarter of its fields being holes.
   The first part of the message is written on a piece of paper through	these
   holes, then the grille is rotated by 90 degrees and the user can
   write down the second part of the message, etc. But this method does only
   work, if the holes are chosen carefully: Every field has to be used, and
   no field may be used twice, either. The ciphertext is read out line by
   line.

@@

%------------------------------------------------------------------------------
% {\bf Column and row transposition}
\subsection[Column and row transposition ciphers]
    {Column and row transposition\footnotemark}
    \footnotetext{%
Most of the following methods can be simulated in CrypTool\index{CrypTool} 
under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
\textbackslash{} Permutation}.}

@@

\item {\bf Simple columnar transposition} \cite{pp:Savard1999}: First of all, 
   a keyword is chosen, that is written above the columns of a table. This
   table is filled with the text to be encrypted line by line. Then the 
   columns are rearranged by sorting the letters of the keyword alphabetically.
   Afterwards the columns are read out from left to right to build the 
   ciphertext\footnote{%
   Using CrypTool: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column.}.

@@

\item {\bf Double column transposition} \cite{pp:Savard1999}
   \index{Double column transposition}: 
   Double columnar transposition
   was frequently used during WW2 and during the Cold War. Two simple columnar
   transpositions with different keys are executed successively\footnote{%
   Using CrypTool: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column. Then choose a (different) key for the
   2nd permutation, input line by line, permute and output column by column.}.\\
   If the keys are different and long enough (at least each 20 characters),
   then this is even for today's computer a real challenge\footnote{%
     MTC3 offers an according level-X challenge called ``Double Columnar
     Transposition'':\\
     \url{http://www.mysterytwisterc3.org/en/crypto-challenges/level-x-cryptography-challenges?id=12}                                                     }.

@@

\item {\bf Row transposition} \cite{pp:Savard1999}: The plaintext is divided 
   into blocks of equal length and a keyword is chosen.
   Now the letters of the keyword are numbered and permutation is done only
   within each block according to this numbering\footnote{%
   Using CrypTool: Choose a key for 1st permutation, input line by line, 
   permute column by column and output line by line.}.

@@

\item {\bf General monoalphabetic substitution / Random letter 
   pairs\footnotemark}
   \footnotetext{%
   This cipher can be simulated in CrypTool\index{CrypTool} 
   under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
   \textbackslash{} Substitution / Atbash}.} 
   \cite{pp:Singh1999}:
   The substitution occurs by a given assignment of single letters.

\item {\bf Atbash\footnotemark}
   \footnotetext{%
   This cipher can be simulated in CrypTool\index{CrypTool} 
   under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
   \textbackslash{} Substitution / Atbash}.}  
   \cite{pp:Singh1999}\index{Atbash}: Replace the first letter of the 
   alphabet by the last letter of the alphabet, the second one by the last 
   but one, etc.

\item {\bf Shift cipher, for example Caesar cipher}\footnote{In CrypTool
   this method  can be found at three different places in the menu tree:\\
   - {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
      Caesar / ROT13}\\
   - {\bf Analysis \textbackslash{} Symmetric Encryption (classic)
     \textbackslash{} Ciphertext only \textbackslash{} Caesar} \\
   - {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Caesar}. } 
   \cite{pp:Singh1999}\index{Caesar}%
   : Plaintext alphabet and ciphertext alphabet are shifted against each 
   other by a determined number of letters. Using the Caesar cipher means 
   shifting letters about three positions.

	Plaintext:	three positions to the right

	Ciphertext:	WKUHH SRVLWLRQV WR WKH ULJKW

\item {\bf Affine cipher}: This is a generalization of the shift
   cipher. A plaintext character is first substituted for another
   character and then the result is encrypted using the shift
   cipher. The affine cipher is so named because its encryption and
   decryption functions are affine or linear functions.

@@

\item {\bf Nihilist Substitution}\footnote{An animation of this Nihilist method
   can be found in CrypTool at the menu item
     {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Nihilist}. }
   \cite{pp:ACA2002}\index{Nihilist substitution}: 
   Insert the alphabet into a 5x5-matrix to assign each letter the number built 
   from row and column number.  
   A keyword is chosen and placed above the columns of a second matrix (grille). 
   The plaintext is written row by row into the grille.
   The ciphertext results from 
   adding the numbers of the plaintext and the numbers of the keyword.
   Numbers between 100 and 110 are transformed to numbers between 00 and
   10, so that each letter is represented by a two-digit number.

@@

\begin{itemize}
\item {\bf Homophonic monoalphabetic substitution}\footnote{This cipher can be 
   simulated in CrypTool\index{CrypTool} under the menu
   {\bf Encrypt/Decrypt \textbackslash{}Symmetric (classic)\textbackslash{} Homophone}.}
   \cite{pp:Singh1999}: Each language has a typical frequency distribution of
   letters. To conceal this distribution, each plaintext letter is assigned
   several ciphertext characters.
   The number of ciphertext characters assigned depends on the frequency of the
   letter to be encrypted.

@@

   \hypertarget{playfair}{}
\item {\bf Playfair}\footnote{%
   In CrypTool\index{CrypTool} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Playfair}.}
   \cite{pp:Singh1999}\index{Playfair}:
   A 5x5-matrix is filled with the plaintext characters. For example, the
   different letters of a keyword are inserted first, followed by the
   remaining letters. The plaintext is divided into pairs, these digraphs
   are encrypted using the following rules:

@@

\begin{itemize}
\item {\bf Vigen\`ere}\footnote{%
   In CT1\index{CT1} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Vigen\`ere}.}  
   \cite{pp:Singh1999}\index{Vigen\`ere}: 
   Each plaintext character is encrypted
   with a different ciphertext alphabet that is determined by the characters of
   a keyword (the so-called Vigen\`ere tableau serves auxiliary means).
   If the plaintext is longer than the key, the latter is repeated.

@@

	\item {\bf Autokey} \cite{pp:Savard1999}: After using the agreed key,
	 use the message itself as a key.
	 See table \ref{Autokey-table-reference}.

@@

       \item {\bf One Time Pad (OTP)}\footnote{%
          CT1\index{CT1} offers this method under the menu
          {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic)
          \textbackslash{} Vernam / OTP}.}\index{One Time Pad}\index{OTP}:
          A sequence
          of bytes is XORed byte-by-byte to the plaintext. This is a
          generalization of Vigen\`ere's mechanism and it was the first
          information-theoretically secure scheme (see
          chapter~\ref{cm_Section_Security_Definitions}
          ``\nameref{cm_Section_Security_Definitions}'').
          To fulfill this claim the pad must be random and it must be used
          only once (to eliminate any semblance of pattern from the ciphertext).\\
          Reason: Given ciphertext C, plaintext P, pad K and two plaintexts
          encrypted with the same key: C1 = P1 $\oplus$ K;  C2 = P2 $\oplus$ K; \\
          Thus, C1 $\oplus$ C2 = (P1 $\oplus$ K) $\oplus$ (P2 $\oplus$ K) = P1 $\oplus$ P2; \\
          which effectively could leak the plaintexts\footnote{%
          In JCT\index{JCrypTool}\index{JCT} you can play with an automatic
          cryptanalysis of running-key ciphertexts using the Viterbi analysis
          under the menu {\bf Visuals \textbackslash{} Viterbi}.}.
   \end{itemize}

@@

\item {\bf ADFG(V)X}\footnote{%
   In CrypTool\index{CrypTool} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
   ADFGVX}.}
   \cite{pp:Singh1999}\index{ADFGVX}%
   : 
   ADFG(V)X-encryption was developed in Germany during WW1. The alphabet 
   is filled into a 5x5 or 6x6 matrix, and columns and rows are marked with
   the letters ADFGX and V, depending on the size of the grille. Each 
   plaintext character is substituted by the corresponding pair of letters. 
   Finally, a (row-) transposition cipher is performed on the resulting text.

@@

\item {\bf Card games} \cite{pp:Savard1999}: The key is created by means of a
   pack of cards and rules that are agreed upon in advance. All methods
   mentioned in this paragraph are designed as paper and pencil methods,
   i.e. they are applicable without electronic aid. A pack of cards is
   unsuspicious to outsiders, shuffling the deck provides a certain amount
   of coincidence, cards can be transformed into numbers easily and a
   transposition cipher can be carried out without any further aid.
   \begin{itemize}
      \item {\bf Solitaire (Bruce Schneier)\footnotemark}
         \footnotetext{%
         %This cipher will be implemented in CrypTool in a future release.}
         In CrypTool\index{CrypTool} you can call this method under the menu
         {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
         Solitaire}.}
         \cite{pp:Schneier1999}\index{Solitaire}: 
         Sender and receiver have to own a deck of cards shuffled in the
         same manner. A key stream is generated that has to consist of as 
         many characters as the message to be encrypted.

@@

\noindent To make the sample code\footnote{%
     Further examples with Sage concerning classic crypto methods can be found e.g.:\\
     - at \url{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}\\
     - in the thesis of Minh Van Nguyen \cite{pp:Nguyen2009}
                                          }
easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are not ASCII characters.
  \end{itemize}
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-en}
\caption{Structure and naming convention of the Sage cipher code examples} 
\label{XXX}
\end{center}
\end{figure}

% ---------------------------------------------------------------------------
\newpage
\subsection{Transposition ciphers}

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitution ciphers}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Caesar cipher}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Shift cipher}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Affine cipher}

@@

In the following example, we construct an affine cipher with key
$(3, 13)$ and use this key to encrypt a given plaintext P. The plaintext
is then decrypted and the result DC is compared to the original plaintext.

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Substitution with symbols}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Vigen{\`e}re cipher}

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Hill cipher}

>>>>

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Writing rule: Capitel header with all nouns/verbs-words capitalized,
%               All other headers in the normal lower/upper case manner
%               like sentences (without a dot at the end).
% Writing rule: Use American English instead of British English consistently.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% Update and corrections: B. Esslinger, June 2005
% Update B. Esslinger and Minh Van Nguyen, November 2009 
% Update B. Esslinger, June 2010 
% Update B. Esslinger, Nov. 2012 
%------------------------------------------------------------------------------

@@

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005;
 Updates Minh Van Nguyen and B. Esslinger, Nov. 2009, June 2010 and Nov. 2012)
\index{Paper and pencil methods}
\index{Encryption!classic}

\begin{ctsquote}
Few persons can be made to believe that it is not quite an easy thing
to invent a method of secret writing which shall baffle investigation.
Yet it may be roundly asserted that human ingenuity cannot concoct a
cipher which human ingenuity cannot resolve.
\caption{Edgar Allan Poe: A Few Words on Secret Writing, 1841}\index{Poe, Edgar Allan}
\end{ctsquote}

The following chapter provides a broad overview of paper and pencil 
methods\footnote{%
The footnotes of this chapter describe how these methods can be
performed using the offline programs CrypTool 1 ({\bf CT1})\index{CT1},
CrypTool 2 ({\bf CT2})\index{CT2} and JCrypTool ({\bf JCT})\index{JCT}.\\
Many of the methods can also per performed within a browser, e.g. on the
the website CrypTool-Online ({\bf CTO}) (\url{http://www.cryptool-online.org})\index{CTO}.
See the appendix \ref{s:appendix-function-overview-CTO}.\\
Additionally the last sub chapter (\ref{PaP_Sage_samples})
contains example code for the computer-algebra system {\bf Sage}\index{Sage}.
                 }
each with references to deeper information.
All techniques that people can apply manually to en- and decipher a message are 
embraced by this term. These methods were and still are especially popular with
secret services, as a writing pad and a pencil -- in contrast to electronic
aids -- are totally unsuspicious.

@@

\item {\bf Rail Fence}\footnote{%
   This method can directly be found in CT1\index{CrypTool 1} at the menu item
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} Scytale / Rail Fence}.
   You can simulate this method also under the menu {\bf Encrypt/Decrypt \textbackslash{} 
   Symmetric (classic) \textbackslash{} Permutation}: For a Rail Fence with
   2 lines use as key ``B,A'' and accept the default settings (only one
   permutation, where your input is done line-by-line and the output is
   taken column-by-column). 
   Using the key ``A,B'' would start the zigzag pattern below in the
   way, that the first letter is written into the first line instead of the
   second line.}
   \cite{pp:Singh1999}\index{Rail Fence cipher}:
   The characters of a message are alternately written in two (or more) lines,
   creating a zigzag pattern. The resulting ciphertext is read out 
   line by line.\\
   This is more a children's method.

   Plaintext\footnote{Agreement: If the alphabet only uses 26 letters, we write from now
   onwards the plaintext in small letters and the ciphertext in capital letters.}%
   : an example of transposition

@@

\item {\bf Scytale}\footnote{%
    This method can directly be found in CT1\index{CrypTool 1} at the menu item
    {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{}
    Scytale / Rail Fence}.
    As this method is a special case of a simple columnar transposition, you also can
    simulate it in CT1\index{CrypTool 1} under the 
    menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
    Permutation}: For the Scytale within the dialog box only the first 
    permutation is used. If the wood has e.g.\ 4 angles use as key ``1,2,3,4''.
    This is equivalent to write the text horizontally in blocks of 4 letters 
    in a matrix and to read it out vertically . 
    Because the key is in an in ascending order, the Scytale is denoted as
    an identical permutation. And because writing and read-out is done only
    once it is a simple (and no double) permutation.\\
    In CT2\index{CrypTool 2} you can find the Scytale within the templates
    {\bf Cryptography \textbackslash{} Classical}).
   }
   \cite{pp:Singh1999}\index{Scytale}%
   : 
   This method was probably used since 600 B.C. -- a description
   of how it operated is not known from before Plutarch (50-120 B.C.).\\
   A long strip of paper is wrapped around a wooden cylinder and then the 
   message is written along the length of this strip. The ciphertext is 
   produced by unwinding the strip.

@@

   \hypertarget{turning-grille}{}
\item {\bf Turning grille} \cite{pp:Savard1999}%
   \index{Turning grille}: 
   The German army used turning 
   grilles during WW1\footnote{%
      The turning grille was already invented in 1881 by Eduard Fleissner von
      Wostrowitz.\\
      A good visualization can be found under www.turning-grille.com.\\
      In JCT\index{JCrypTool}\index{JCT} you can find it in the default perspective
      via the menu item {\bf Visuals \textbackslash{} Grille}.
   }%
   . 
   A square grille serves as a stencil, a quarter of its fields being holes.
   The first part of the message is written on a piece of paper through	these
   holes, then the grille is rotated by 90 degrees and the user can
   write down the second part of the message, etc. But this method does only
   work, if the holes are chosen carefully: Every field has to be used, and
   no field may be used twice, either. The ciphertext is read out line by
   line.

@@

%------------------------------------------------------------------------------
% {\bf Column and row transposition}
\subsection[Column and row transposition ciphers]
    {Column and row transposition\footnotemark}
    \footnotetext{%
Most of the following methods can be simulated in CT1\index{CrypTool 1} 
under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
\textbackslash{} Permutation}.}

@@

\item {\bf Simple columnar transposition} \cite{pp:Savard1999}: First of all, 
   a keyword is chosen, that is written above the columns of a table. This
   table is filled with the text to be encrypted line by line. Then the 
   columns are rearranged by sorting the letters of the keyword alphabetically.
   Afterwards the columns are read out from left to right to build the 
   ciphertext\footnote{%
   Using CT1: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column.}.

@@

\item {\bf Double column transposition} \cite{pp:Savard1999}
   \index{Double column transposition}: 
   Double columnar transposition
   was frequently used during WW2 and during the Cold War. Two simple columnar
   transpositions with different keys are executed successively\footnote{%
   Using CT1: Choose a key for the 1st permutation, input line by line, 
   permute and output column by column. Then choose a (different) key for the
   2nd permutation, input line by line, permute and output column by column.}.\\
   If the keys are different and long enough (at least each 20 characters),
   then this is even for today's computer a real challenge\footnote{%
     MTC3\index{MTC3} offers an according level-X challenge called ``Double Columnar
     Transposition'':\\
     \url{http://www.mysterytwisterc3.org/en/challenges/level-x/double-column-transposition}
   }.

@@

\item {\bf Row transposition} \cite{pp:Savard1999}: The plaintext is divided 
   into blocks of equal length and a keyword is chosen.
   Now the letters of the keyword are numbered and permutation is done only
   within each block according to this numbering\footnote{%
   Using CT1: Choose a key for 1st permutation, input line by line, 
   permute column by column and output line by line.}.

@@

\item {\bf General monoalphabetic substitution / Random letter 
   pairs\footnotemark}
   \footnotetext{%
    This cipher can be simulated in CT1\index{CrypTool 1} 
    under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
    \textbackslash{} Substitution / Atbash}.\\
    In CT2\index{CrypTool 2} you can find these methods within the templates
    {\bf Cryptography \textbackslash{} Classical}).
   } 
   \cite{pp:Singh1999}:
   The substitution occurs by a given assignment of single letters.

\item {\bf Atbash\footnotemark}
   \footnotetext{%
   This cipher can be simulated in CT1\index{CrypTool 1} 
   under the menu {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) 
   \textbackslash{} Substitution / Atbash}.}  
   \cite{pp:Singh1999}\index{Atbash}: Replace the first letter of the 
   alphabet by the last letter of the alphabet, the second one by the last 
   but one, etc.

\item {\bf Shift cipher, for example Caesar cipher}\footnote{In CT1\index{CrypTool 1}
   this method  can be found at three different places in the menu tree:\\
   - {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
      Caesar / ROT13}\\
   - {\bf Analysis \textbackslash{} Symmetric Encryption (classic)
     \textbackslash{} Ciphertext only \textbackslash{} Caesar} \\
   - {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Caesar}. } 
   \cite{pp:Singh1999}\index{Caesar}%
   : Plaintext alphabet and ciphertext alphabet are shifted against each 
   other by a determined number of letters. Using the Caesar cipher means 
   shifting letters about three positions.

    Plaintext:  \texttt{		three positions to the right}

    Ciphertext: \texttt{		WKUHH SRVLWLRQV WR WKH ULJKW}

\item {\bf Affine cipher}\footnote{
   Some according Sage samples are implemented at \ref{PaP_affine_cipher_sage-sample}.
   }%
   \index{Affine cipher}: This is a generalization of
   the shift cipher. A plaintext character is first substituted by another
   character and then the result is encrypted using the shift cipher.
   The name ``affine cipher'' was chosen because its encryption and
   decryption can be described as affine or linear function.

@@

\item {\bf Nihilist Substitution}\footnote{An animation of this Nihilist method
   can be found in CT1\index{CrypTool 1} at the menu item
     {\bf Indiv. Procedures \textbackslash{} Visualization of Algorithms 
     \textbackslash{} Nihilist}. }
   \cite{pp:ACA2002}\index{Nihilist substitution}: 
   Insert the alphabet into a 5x5-matrix to assign each letter the number built 
   from row and column number.  
   A keyword is chosen and placed above the columns of a second matrix (grille). 
   The plaintext is written row by row into the grille.
   The ciphertext results from 
   adding the numbers of the plaintext and the numbers of the keyword.
   Numbers between 100 and 110 are transformed to numbers between 00 and
   10, so that each letter is represented by a two-digit number.

@@

\begin{itemize}
\item {\bf Homophonic monoalphabetic substitution}\footnote{This cipher can be 
   simulated in CT1\index{CrypTool 1} under the menu
   {\bf Encrypt/Decrypt \textbackslash{}Symmetric (classic)\textbackslash{} Homophone}.}
   \cite{pp:Singh1999}: Each language has a typical frequency distribution of
   letters. To conceal this distribution, each plaintext letter is assigned
   several ciphertext characters.
   The number of ciphertext characters assigned depends on the frequency of the
   letter to be encrypted.

@@

   \hypertarget{playfair}{}
\item {\bf Playfair}\footnote{%
   In CT1\index{CrypTool 1} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic)
   \textbackslash{} Playfair}.\\
   In JCT\index{JCrypTool}\index{JCT} you can find it in the default perspective
   via the menu item {\bf Algorithms \textbackslash{} Classic \textbackslash{}
   Playfair}.}
   \cite{pp:Singh1999}\index{Playfair}:
   A 5x5-matrix is filled with the plaintext characters. For example, the
   different letters of a keyword are inserted first, followed by the
   remaining letters. The plaintext is divided into pairs, these digraphs
   are encrypted using the following rules:

@@

\begin{itemize}
\item {\bf Vigen\`ere}\footnote{%
   In CT1\index{CT1} you can call this method under the menu {\bf Encrypt/Decrypt
   \textbackslash{} Symmetric (classic) \textbackslash{} Vigen\`ere}.\\
   In CT2\index{CrypTool 2} you can find this method within the templates
   {\bf Cryptography \textbackslash{} Classical}.\\
   In JCT\index{JCrypTool}\index{JCT} you can find it in the default perspective
   via the menu item {\bf Algorithms \textbackslash{} Classic \textbackslash{}
   Vigen\`ere}.
   }  
   \cite{pp:Singh1999}\index{Vigen\`ere}: 
   Each plaintext character is encrypted
   with a different ciphertext alphabet that is determined by the characters of
   a keyword (the so-called Vigen\`ere tableau serves auxiliary means).
   If the plaintext is longer than the key, the latter is repeated.

@@

	\item {\bf Autokey}\footnote{%
   In CT2\index{CrypTool 2} you can find this method within the templates
   {\bf Cryptography \textbackslash{} Classical}.\\
   In JCT\index{JCrypTool}\index{JCT} you can find it in the default perspective
   via the menu item {\bf Algorithms \textbackslash{} Classic \textbackslash{}
   Autokey Vigen\`ere}.
   } \cite{pp:Savard1999}: After using the agreed key,
	 use the message itself as a key.
	 See table \ref{Autokey-table-reference}.

@@

       \item {\bf One Time Pad (OTP)}\footnote{%
          CT1\index{CT1} offers this method under the menu
          {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic)
          \textbackslash{} Vernam / OTP}.\\
          In CT2\index{CrypTool 2} you can find this method within the templates
          {\bf Cryptography \textbackslash{} Classical}.\\
          In JCT\index{JCrypTool}\index{JCT} you can find it in the default perspective
          via the menu item {\bf Algorithms \textbackslash{} Classic
          \textbackslash{} XOR}.
          }\index{One Time Pad}\index{OTP}:
          A sequence
          of bytes is XORed byte-by-byte to the plaintext. This is a
          generalization of Vigen\`ere's mechanism and it was the first
          information-theoretically secure scheme (see
          chapter~\ref{cm_Section_Security_Definitions}
          ``\nameref{cm_Section_Security_Definitions}'').
          To fulfill this claim the pad must be random and it must be used
          only once (to eliminate any semblance of pattern from the ciphertext).\\
          Reason: Given ciphertext C, plaintext P, pad K and two plaintexts
          encrypted with the same key: C1 = P1 $\oplus$ K;  C2 = P2 $\oplus$ K; \\
          Thus, C1 $\oplus$ C2 = (P1 $\oplus$ K) $\oplus$ (P2 $\oplus$ K) = P1 $\oplus$ P2; \\
          which effectively could leak the plaintexts\footnote{%
          In JCT\index{JCrypTool}\index{JCT} you can play with an automatic
          cryptanalysis of running-key ciphertexts using the Viterbi analysis
          under the menu {\bf Visuals \textbackslash{} Viterbi}.}.
   \end{itemize}

@@

\item {\bf ADFG(V)X}\footnote{%
   In CT1\index{CrypTool 1} you can call this method under the menu
   {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
   ADFGVX}.\\
   In CT2\index{CrypTool 2} you can find this method within the templates
   {\bf Cryptography \textbackslash{} Classical}.
   }
   \cite{pp:Singh1999}\index{ADFGVX}%
   : 
   ADFG(V)X-encryption was developed in Germany during WW1. The alphabet 
   is filled into a 5x5 or 6x6 matrix, and columns and rows are marked with
   the letters ADFGX and V, depending on the size of the grille. Each 
   plaintext character is substituted by the corresponding pair of letters. 
   Finally, a (row-) transposition cipher is performed on the resulting text.

@@

\item {\bf Card games} \cite{pp:Savard1999}: The key is created by means of a
   pack of cards and rules that are agreed upon in advance. All methods
   mentioned in this paragraph are designed as paper and pencil methods,
   i.e. they are applicable without electronic aid. A pack of cards is
   unsuspicious to outsiders, shuffling the deck provides a certain amount
   of coincidence, cards can be transformed into numbers easily and a
   transposition cipher can be carried out without any further aid.
   \begin{itemize}
      \item {\bf Solitaire (Bruce Schneier)\footnotemark}
         \footnotetext{%
         In CT1\index{CrypTool 1} you can call this method under the menu
         {\bf Encrypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
         Solitaire}.}
         \cite{pp:Schneier1999}\index{Solitaire}: 
         Sender and receiver have to own a deck of cards shuffled in the
         same manner. A key stream is generated that has to consist of as 
         many characters as the message to be encrypted.

@@

\noindent To make the sample code\footnote{%
     Further examples with Sage concerning classic crypto methods can be found e.g.:\\
     - at \url{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}\\
     - in the thesis of Minh Van Nguyen \cite{pp:Nguyen2009}
                                          }
of the ciphers easier to understand, we used the structure and the naming conventions
shown in the Figure \ref{naming-convention-in-Sage-cipher-samples} below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are not ASCII characters.
  \end{itemize}
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.5]{figures/notation-enc-dec_in-sage-scripts_en}
\caption{Structure and naming convention of the Sage cipher code examples} 
\label{naming-convention-in-Sage-cipher-samples}
\end{center}
\end{figure}

% ---------------------------------------------------------------------------
\newpage
\subsection{Transposition ciphers}
\index{Transposition}

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Substitution ciphers}
\index{Substitution}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Caesar cipher}
\index{Caesar}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Shift cipher}
\index{Shift cipher}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Affine cipher}
\label{PaP_affine_cipher_sage-sample}
\index{Affine cipher}

@@

In the following example, we construct an affine cipher $ c_i = b*p_i + a $
with key $(3, 13)$ and use this key to encrypt a given plaintext
$P = (p_1,p_2, ...,p_n)$. The plaintext is then decrypted and the result
DC is compared to the original plaintext.

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Substitution with symbols}
\index{Substitution}

@@

% ---------------------------------------------------------------------------
\newpage
\subsubsection{Vigen{\`e}re cipher}
\index{Vigen\`ere}

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Hill cipher}
\index{Hill}
